\propertitle

### 2019-11-13 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-13T16:55:17+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{p_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycmm}{C}
\newcommand*{\ycm}{\mathte{\ycmm}}
\newcommand*{\ycs}{\bm{\sigma}}
\newcommand*{\yL}{L}
\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\yds}{\delta}

@@

The model we propose is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

@@

Regarding~\ref{item:maxent_step}, we assume familiarity with the
maximum-entropy method. We actually use the \emph{minimum-relative-entropy}
method \citep{hobsonetal1973}, but call it \enquote{maximum-entropy} for
brevity. It amounts to a pair of prescriptions: choose the distribution,
among those satisfying specific convex constraints, such as fixed
expectations, that has minimum relative entropy with respect to a reference
distribution, often taken to be the uniform one; and judge those
expectations to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First remark: the
distribution $\px(\yA)$ given by this method is the zeroth-order
approximation \citep[in the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{itemize}
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{itemize}
The maximum-entropy distribution is thus an approximation of our belief
distribution about four different quantities. Note that the four
distributions above numerically differ in higher-order approximations.
Second remark: the maximum-entropy method based on the Shannon entropy
implicitly makes some assumptions about the probabilities for the long-run
frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these two
points further in \sect~\ref{sec:assumptions}.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{\ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya=0}^{n} %m!\,
  \displaystyle\binom{\ya}{m} \bigg/ \binom{n}{m}\; \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $a=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Fixing the ones is therefore equivalent
to fixing the others. But the normalized factorial moments have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{\ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{\yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is our distribution of belief about the full-network
activity.

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network by constraining its $m$th factorial moment to be equal
to the sample's recorded average of $\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}$,
for as many $m$ as we please with $1\le m \le n$. In formulae these
constraints on $\px(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{\ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}}_{\text{measured moments}}
  =
   \underbracket{\sum_{\yA}
  \binom{\yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)}_{\text{distribution moments}}
\end{equation}
for all $m$ we wish.

The calculation amounts to a convex optimization
\citep{meadetal1984}[\chap~10]{pressetal1988_r2007}{fangetal1997,boydetal2004_r2009,portamana2017b}
and for the numbers $N,n,m$ considered in the present work it can be done
on a modern computer without approximations of normalization constants or
potential functions.

@@

For concreteness's sake we'll consider the maximum-entropy distribution
$\px(\yA)$ as \emph{the most probable frequency distribution} for the
full-network activity during the recording; but remember that it is also
the approximation of three other distributions, as discussed in the
previous section.

@@

We first calculate the distribution by using six moments. This number
already provides almost as much information as the full frequency
distribution of the sample, and at the same time illustrates the use of the
approach in questions of statistic sufficiency (typically limited to two or
three moments). Figure~\ref{fig:plots_6mom} shows the resulting densities
(that is, distribution${}\times N$) for full-network sizes $N=n$,
$N=1\,000$, $N=5\,000$, $N=10\,000$\mynote{motivate?}. The case $N=n$
corresponds to applying the maximum-entropy method at the sample level; it
can be observed that with six moments it reproduces almost exactly the
measured frequency distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-13T19:14:02+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

\newcommand*{\haleprint}[1]{%\global\def\arxivp{\arxivsi}%\citeauthor{0arxivcite}\addtocategory{ifarchcit}{0arxivcite}%eprint
\texttt{\urlalt{https://hal.archives-ouvertes.fr/#1}{HAL:\hspace{0pt}#1}}%

@@

\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}
% https://tex.stackexchange.com/a/33547/97039
%\newcommand*{\fm}{f}
\newcommand*{\px}{P_{\text{me}}}
\newcommand*{\pxx}{p}
\newcommand*{\pxxx}{P}
\newcommand*{\Nd}{\textrm{N}}
\newcommand*{\ycmm}{C}
\newcommand*{\ycm}{\mathte{\ycmm}}
\newcommand*{\ycs}{\bm{\sigma}}
\newcommand*{\yL}{L}
\newcommand*{\yd}{\bm{\varDelta}}
\newcommand*{\yds}{\delta}

@@

The model we propose is based on the straightforward combination of the
maximum-entropy method and basic sampling relations from the probability
calculus, discussed in \sect~\ref{sec:model}. The maximum-entropy or
minimum-relative-entropy method \citep{jaynes1957}[much clearer
in][]{jaynes1963}{sivia1996_r2006,hobsonetal1973,jaynes1985b_r1996,grandy1980}
has been used for different kinds of estimations of the neuronal activity
of various brain areas and about other phenomena of importance to the
neurosciences, for example gene and protein interaction \citep[for
example][]{martignonetal1995,bohteetal2000,shlensetal2006,schneidmanetal2006,tkaciketal2006,mackeetal2009b,tkaciketal2009,roudietal2009c,barreiroetal2010,gerwinnetal2010,mackeetal2011b,ganmoretal2011,cohenetal2011,granotatedgietal2013,mackeetal2013,tkaciketal2014b,shimazakietal2015,moraetal2015,lezonetal2006,weigtetal2009}.

@@

Regarding~\ref{item:maxent_step}, we assume familiarity with the
maximum-entropy method. We actually use the \emph{minimum-relative-entropy}
method \citep{hobsonetal1973}, but call it \enquote{maximum-entropy} for
brevity. It amounts to a pair of prescriptions: choose the distribution,
among those satisfying specific convex constraints, such as fixed
expectations, that has minimum relative entropy with respect to a reference
distribution, often taken to be the uniform one; and judge those
expectations to be equal to measured averages. We add two remarks about
this method that are seldom made in the literature. First remark: the
distribution $\px(\yA)$ given by this method is the zeroth-order
approximation \citep[in the sense of Laplace's
method:][\chap~4]{debruijn1958_r1961}{tierneyetal1986,strawderman2000} of
four different distributions for the full network:
\begin{enumerate}[label=(\roman*)]
\item the most probable \emph{frequency} distribution for the total
  activity across the \emph{recorded} bins, %$\yFF(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
  %$\yp(\yA_{t} \| \yH)$ with $1\le t \le T$,
\item the most probable \emph{frequency} distribution for the total
  activity in a very long run of \emph{new} time bins, %$\ynuu(\yA)$,
\item the \emph{belief} distribution for the value of the total
  activity at a \emph{new} time bin.
  % $\yp(\yA_{t} \| \yH)$ with $t<1$ or $t>T$
\end{enumerate}
The maximum-entropy distribution is thus an approximation of our belief
distribution about four different quantities. Note that the four
distributions above numerically differ in higher-order approximations.
Second remark: the maximum-entropy method based on the Shannon entropy
implicitly makes some assumptions about the probabilities for the long-run
frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these two
remarks further in \sect~\ref{sec:assumptions}.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{\ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya=0}^{n} %m!\,
  \displaystyle\binom{\ya}{m} \bigg/ \binom{n}{m}\; \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $a=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Specifying one set is therefore
equivalent to specifying the other set. But the normalized factorial
moments have a convenient property for our analysis
\citep[\sect~2.3]{portamanaetal2015}{potts1953}: \emph{the first $n$
  normalized factorial moments for the sample and for the full network are
  numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya=0}^{n} %m!\,
  \binom{\ya}{m}\bigg/  \binom{n}{m}\; \pxx(\ya)
=    \sum_{\yA=0}^{N} %m!\,
  \binom{\yA}{m}\bigg/  \binom{N}{m}\; \pxxx(\yA),
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxxx(\yA)$ is our distribution of belief about the full-network
activity.

We can therefore apply the maximum-entropy method to obtain a distribution
$\px(\yA)$ for the full network by constraining its $m$th factorial moment
to be equal to the sample's recorded average of
$\tbinom{\ya}{m}/\tbinom{n}{m}$, for as many $m$ as we please with
$1\le m \le n$. In formulae, the constraints on $\px(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{\ya_{t}}{m}\bigg/\binom{n}{m}
  \equiv
   \sum_{\ya}
  \binom{\ya}{m}\bigg/  \binom{n}{m} \; \yff_{\ya}}_{\text{measured moments}}
  \mathrel{\;=\;}
   \underbracket{\sum_{\yA}
  \binom{\yA}{m}\bigg/  \binom{N}{m} \; \px(\yA)}_{\text{distribution moments}}.
\end{equation}

Finding the constrained maximum-entropy distribution $\px(\yA)$ amounts to
a convex optimization
\citep{meadetal1984}[\chap~10]{pressetal1988_r2007}{fangetal1997,boydetal2004_r2009,portamana2017b}
and for the numbers $N,n,m$ considered in the present work it can be done
on a modern computer without approximations of the partition functions that
appear in the formalism.

@@

% In the followingFrom a Bayesian point of view, the maximum-entropy distribution $\px(\yA)$
% can be considered as \emph{the most probable frequency distribution} for
% the full-network activity during the recording. But it also approximates
% three other distributions, as discussed in the previous section.

@@

We first calculate the distribution by using six moments. This number
already provides almost as much information as the full frequency
distribution of the sample, and at the same time illustrates the use of the
approach in questions of statistical sufficiency \citep[cf. for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
Figure~\ref{fig:plots_6mom} shows the resulting densities (that is,
distribution${}\times N$) for full-network sizes $N=n$, $N=1\,000$,
$N=5\,000$, $N=10\,000$. The case $N=n$ corresponds to applying the
maximum-entropy method at the sample level; it can be observed that with
six moments it reproduces almost exactly the measured frequency
distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb
